Local Spectral Clustering of Density Upper Level Sets

### 2019-10-04 arxiv/examples_section.tex
added relevant examples tuff

\begin{abstract}
\noindent We analyze the Personalized PageRank (PPR) algorithm, a local spectral method for clustering,
which extracts clusters using locally-biased random walks around a user-specified
seed node.  In contrast to previous work, we adopt a traditional statistical
learning setup, where we obtain samples from an unknown distribution, and aim to
identify connected regions of high-density (density clusters).  We prove that
PPR, run on a neighborhood graph, extracts sufficiently salient density
clusters. We also provide empirical support for our theory.
\end{abstract}
\end{center}
\section{Introduction}
\label{sec: introduction}
In this paper, we consider the problem of clustering: splitting a given data set
into groups that satisfy some notion of within-group similarity and
between-group difference.  Our particular focus is on
spectral clustering methods, a family of powerful
nonparametric clustering algorithms. Roughly speaking, a spectral algorithm
first constructs a geometric graph $G$, where vertices correspond to samples,
and edges correspond to proximities between samples. The algorithm 
then estimates a feature
embedding based on (an appropriate) Laplacian matrix of the graph 
$G$, and applies a simple clustering
technique (like $k$-means clustering) in the embedded feature space.

% Spectral clustering and density clustering
%Spectral clustering algorithms, which
%originated in the early work of Fiedler \cite{fiedler73}, have seen a recent revival 
%(see for instance the recent review~\cite{XX}) focused mainly on understanding the performance of spectral clustering on the Stochastic Block Model (SBM) and closely related problems.
%A parallel line of work \cite{shi09,schiebinger15}, has considered spectral partitioning applied to a non-parametric mixture model
%where once again the clustering goal is to recover the latent labels of a sample $X_1,\ldots,X_n \sim P$. These works proceed by relating the eigenvectors of the graph Laplacian to eigenfunctions of an appropriately defined integral operator, and then using assumptions about the (non-)overlap of the mixture components to show that these eigenfunctions elucidate the latent labels.

When applied to geometric graphs built from a large number of samples,
global spectral clustering methods can be computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to increased interest in
local spectral clustering algorithms, which leverage locally-biased spectra
computed using random walks around some user-specified seed node.  A popular 
local clustering algorithm is the Personalized PageRank (PPR) algorithm, first introduced by 
\citet{haveliwala2003}, then further developed by
several others~\citep{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013}.

Local spectral clustering techniques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, leading 
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (such as
conductance).  In this paper, we adopt a classical statistical viewpoint,
and examine what the output of local clustering on a data set reveals about the
underlying density of the samples 
$f$.  In particular, we examine the ability of PPR to recover
\emph{density clusters} of $f$, defined as the connected components of
the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$ for some $\lambda > 0$
(a central object of interest in the statistical clustering literature, dating
back to the work of \citet{hartigan1981}).

\section{Background and Related Work}
We begin by providing some standard background on the PPR algorithm
and the density clustering setup, before turning our attention to related work
and a
detailed summary of our contributions.

\subsection{PPR on a neighborhood graph} 
We now describe the clustering
algorithm that will be our focus for the rest of the paper. Let $\Xbf = \{x_1,
\ldots, x_n\}$ be a sample drawn i.i.d.\ from a distribution $\Pbb$ on $\Rd$,
with density~$f$.  For a radius $r > 0$, we define $G_{n,r}=(V,E)$ to be the
\emph{$r$-neighborhood graph} of $\Xbf$, an unweighted, undirected graph with
vertices $V=\Xbf$, and an edge $(x_i,x_j) \in E$ if and only if $\norm{x_i -
x_j} \leq r$, where $\norm{\cdot}$ is the $\ell_2$ norm. We denote by $\Abf \in
\Reals^{n \times n}$ the adjacency matrix, with entries $\Abf_{uv} = 1$ if
$(u,v) \in E$ and $0$ otherwise.  We also denote by $\Dbf$ the diagonal degree
matrix, with $\Dbf_{uu} = \sum_{v \in V} \Abf_{uv}$, and by $\Ibf{}$ the $n
\times n$ identity matrix.

Next, we define the PPR vector $\pbf = \pbf(v,\alpha;G_{n,r})$, based on
a seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to be 
the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf = (\Ibf{} + \Dbf^{-1}\Abf)/2$ is the lazy random walk matrix over
$G_{n,r}$ and $e_{v}$ is the indicator vector for node $v$ (that has a 1 in the
$v$-th position and 0 elsewhere). For a level $\beta > 0$ and a target volume
$\vol_0 > 0$, we define a \emph{$\beta$-sweep cut} of $\pbf = (p_u)_{u \in V}$
as   
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta := \set{u \in V: \frac{p_u}{\Dbf_{uu}} > \frac{\beta}{\vol_{0}}}.
\end{equation}
We will use the normalized cut metric to determine which sweep cut $S_{\beta}$
is the best cluster estimate. For a set $S \subseteq V$ with complement $S^c = V
\setminus S$, we define \smash{$\cut(S;G_{n,r}) := \sum_{u \in S, v \in S^c}
\Abf_{uv}$}, and \smash{$\vol(S; G_{n,r}) := \sum_{u \in S} \Dbf_{uu}$}.  We
define the \emph{normalized cut} of $S$ as
\begin{equation}
\label{eqn: normalized_cut}
\Phi(S; G_{n,r}) := \frac{\cut(S;G_{n,r})}{\min \set{\vol(S; G_{n,r}), \vol(S^c; G_{n,r})}}.
\end{equation}
Having computed sweep cuts $S_{\beta}$ over a range \smash{$\beta \in  
(\frac{1}{40},\frac{1}{11})$}\footnote{The choice of a specific range such as  
\smash{$(\frac{1}{40}, \frac{1}{11})$} is standard in the analysis of PPR
algorithms, see, e.g., \citep{zhu2013}.}, we output the cluster estimate
\smash{$\Cest = S_{\beta^*}$} that has minimum normalized cut.
%\smash{$\Phi(S_{\beta^*}; G_{n,r})$}. 
For concreteness, the PPR algorithm is summarized in Algorithm~\ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a neighborhood graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, target stationary volume
$\vol_0 > 0$. \\     
{\bfseries Output:} cluster $\Cest \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $p=\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn: 
    ppr_vector}. 
  \STATE For \smash{$\beta \in (\frac{1}{40}, \frac{1}{11})$} compute sweep cuts 
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return the cluster \smash{$\Cest = S_{\beta^*}$}, where  
  $$
  \beta^* = \argmin_{\beta \in (\frac{1}{40}, \frac{1}{11})}~ \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

\subsection{Estimation of density clusters} 
Let \smash{$\Cbb_f(\lambda)$} denote 
the connected components of the density upper level set $\{x \in \Rd: f(x) >
\lambda\}$.  For a given density cluster \smash{$\Cset \in \Cbb_f(\lambda)$}, we
call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical density cluster}. The
size of the symmetric set difference between estimated and empirical clusters is 
a commonly used metric to quantify cluster estimation error
\citep{korostelev1993,polonik1995,rigollet2009}.

\begin{definition}[Symmetric set difference]
  \label{def: symmetric_set_diff}
  For an estimator \smash{$\Cest \subseteq \Xbf$} and set
  $\mathcal{S} \subseteq \Reals^d$, we define   
  \begin{equation}
    \label{eqn: misclassification_rate}
    \Delta(\Cest, \mathcal{S}) := \abs{\Cest \setminus \mathcal{S}[\Xbf] \cup
      \mathcal{S}[\Xbf] \setminus \Cest},
  \end{equation}
  the cardinality of the symmetric set difference between 
  \smash{$\Cest$} and $\mathcal{S} \cap \Xbf = \mathcal{S}[\Xbf]$. 
\end{definition}

However, the symmetric set difference does not measure whether \smash{$\Cest$} 
can distinguish any two distinct clusters \smash{$\Cset,\Cset' \in
  \Cbb_f(\lambda)$}. We therefore also study a second notion of cluster
estimation, first introduced by \citet{hartigan1981}, and defined
asymptotically.

\begin{definition}[Consistent density cluster estimation]
  \label{def: consistent_density_cluster_estimation}
  For an estimator \smash{$\Cest \subseteq \Xbf$} and cluster 
  \smash{$\Cset \in \Cbb_f(\lambda)$}, we say \smash{$\Cest$} is a consistent 
  estimator of $\Cset$ if for all \smash{$\Cset' \in \Cbb_f(\lambda)$} with
  $\Cset \not= \Cset'$, the following holds as $n \to \infty$: 
  \begin{equation}
    \label{eqn: consistent_density_cluster_recovery}
    \Cset[\Xbf] \subseteq \Cest \quad \text{and} \quad
    \Cest \cap \Cset'[\Xbf] = \emptyset,
  \end{equation}
  with probability tending to 1.
\end{definition}
\noindent Consistent cluster recovery roughly ensures that, for a given $\lambda$, the estimated cluster $\Cest$ contains all
points in a true
density cluster $\Cset$, and simultaneously does not contain any points in any other density cluster $\Cset'$.

With these definitions in place, our broad goal will be to understand the extent to which the PPR algorithm is able to
recover a cluster which either guarantees low symmetric set difference to a true density cluster, or which consistently estimates
a true density cluster.

%\subsection{Our Contributions}

\subsection{Related Work and Our Contributions}
In addition to the background on local spectral
clustering given previously, a few related lines of work are worth
highlighting. \citet{shi2009} and \citet{schiebinger2015} examine the consistency of  
spectral algorithms in recovering the latent labels in certain
nonparametric mixture models. Their results focus on global rather than local 
methods, and thus impose global rather than local conditions on the nature
of the density. Moreover, they do not in general guarantee recovery of density 
clusters, which is the focus in our work. Perhaps most importantly, these works
rely on general cluster saliency conditions, which implicitly depend on many
distinct geometric aspects of the cluster $\Cset$ under consideration. We make
this dependence more explicit, and in doing so expose the role each geometric
condition plays in the clustering problem.

More broadly, density clustering and level set estimation is a well-studied
problem. \citet{polonik1995, rigollet2009} study density clustering under the 
symmetric set difference metric, \citet{tsybakov1997, singh2009} describe
minimax optimal level-set estimators under Hausdorff loss and
\citet{hartigan1981, chaudhuri2010, balakrishnan2013} consider consistent estimation of the
cluster tree.
%, to note but a few works. 
We emphasize that our goal is not to improve on these
results, nor to offer a better algorithm for level set estimation; indeed, seen as
a density clustering algorithm, PPR has none of the optimality guarantees 
found in the aforementioned works. Instead, our motivation is to start with a 
widely-used local spectral method, PPR, and to better understand and
characterize the distinctions between those density clusters which are
well-conditioned for PPR, and those which are not.

A summary of our results (and an outline for this
paper) is as follows.

\begin{enumerate}
\item In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we introduce
  a set of natural geometric conditions on the density cluster $\Cset$
  %formalize a measure of difficulty based on these geometric conditions, 
  and show that when Algorithm \ref{alg: ppr} is properly initialized, the size of
  the symmetric set difference of \smash{$\Cest$} and a thickened version of the
  density cluster $\Csig$ can be bounded in a meaningful way.
  % this difficulty measure.

\item We further show in Section \ref{sec:
    consistent_cluster_estimation_with_ppr} that if the density cluster 
  $\Cset$ is particularly well-conditioned, Algorithm \ref{alg: ppr}
  will consistently estimate a density cluster in the sense of
  \eqref{eqn: consistent_density_cluster_recovery}.

\item In Section \ref{sec: analysis}, we detail some of the analysis required to
  prove our main results, and expose the parts various geometric quantities play 
  in the difficulty of the clustering problem.

\item In Section \ref{sec: experiments}, we empirically investigate the
  tightness of our analysis, and provide examples showing how violations of our
  geometric conditions impact density cluster recovery by PPR.
\end{enumerate}

\noindent At this point we summarize one of our main takeaways: 
PPR, run on a neighborhood graph, recovers \emph{geometrically
compact} high-density clusters. Our main theoretical results, make this takeaway 
precise, and provide concrete ways of quantifying the geometric
compactness of a density cluster. To complement this, we also provide experimental
evidence that the PPR algorithm can fail to recover high-density clusters which are not
geometrically compact.

\section{Main Results}
\label{sec: consistent_cluster_estimation_with_ppr}
In this section we present our main results on the accuracy of the PPR algorithm 
for recovering density clusters. 
We begin by formally introducing a variety of geometric conditions, 
and use these to define a condition
number $\kappa(\Cset)$, which measures the difficulty PPR will have in  
estimating a density cluster $\Cset$.
With this condition in place 
our first main result (Theorem~\ref{thm: misclassification_rate}) provides a bound
on the symmetric set difference between the estimated cluster $\Cest$,
obtained by an appropriately initialized version of the PPR algorithm,
and
the target cluster $\Cset$.
Our next main result (Theorem~\ref{thm: consistent_recovery_of_density_clusters})
shows that for sufficiently well-conditioned target clusters $\Cset$, the output 
of the PPR algorithm $\Cest$
is consistent in the sense defined in Definition~\ref{def: consistent_density_cluster_estimation}.

\subsection{Preliminaries}%: Geometric Conditions and the Condition Number of a Density Cluster}
%We formalize some geometric conditions, and use these to define a condition
%number $\kappa(\Cset)$, which measures the difficulty PPR will have in  
%estimating $\Cset$. (Our theoretical guarantees for PPR will be framed in terms 
%of $\kappa(\Cset)$.)
%\paragraph{Geometric conditions on density clusters.} 
At a high level, for PPR
to be successful, the underlying density cluster must be geometrically
well-conditioned.  A basic requirement is that we need to avoid clusters which contain arbitrarily
thin bridges or spikes. As in the work of \citet{chaudhuri2010}, we consider a
thickened version of \smash{$\Cset \in \Cbb_f(\lambda)$} defined as 
$\Csig := \set{x \in \Reals^d: \dist(x,\Cset) \leq \sigma}$, which 
we call the \emph{$\sigma$-expansion} of $\Cset$. Here 
\smash{$\dist(x,\Cset) := \inf_{y \in \Cset} \norm{y - x}$}.  We now list our
conditions on $\Csig$.

\begin{enumerate}[label=(A\arabic*)]
\item
  \label{asmp: bounded_density}
  \emph{Bounded density within cluster:} There exist constants
  $0<\lambda_{\sigma}< \Lambda_{\sigma}<\infty$ such that 
  $\lambda_{\sigma} \leq \inf_{x \in \Csig} f(x) \leq \sup_{x \in \Csig} f(x)
  \leq \Lambda_{\sigma}$.

\item
  \label{asmp: cluster_separation}
  \emph{Cluster separation:}
  For all clusters $\Cset' \in \Cbb_f(\lambda)$ with $\Cset' \not= \Cset$, 
  $\dist(\Csig,\Csig') > \sigma$, where $\dist(\Csig,\Csig') := \inf_{x
    \in \Csig} \dist(x,\Csig')$.

\item 
  \label{asmp: low_noise_density}
  \emph{Low noise density:} There exist $\gamma,c_0 > 0$ such that for 
  $x \in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
  \begin{align*}
  \inf_{x' \in \Csig} f(x') - f(x) \geq  c_0 \dist(x, \Csig)^{\gamma}. 
  \end{align*}
  Roughly, 
  this assumption ensures that the density decays sufficiently quickly as we move away from
  the target cluster $\Csig$, and is a standard assumption in the level-set estimation literature
  (see for instance~\cite{singh2009}).

  %%% AJG 5/20: Should I turn this into two conditions?
\item
  \label{asmp: embedding}
  \emph{Lipschitz embedding:}
  There exists $g: \Reals^d \to \Reals^d$ with the following properties: 
  \begin{enumerate}
  \item we have $\Csig = g(\mathcal{K})$, for a convex set $\mathcal{K} \subseteq \Rd$
  with $\mathrm{diam}(\mathcal{K}) = \sup_{x,y \in \mathcal{K}}\norm{x - y} =:
  \diam < \infty$
  \item $\det(\nabla g (x)) = 1$ for all $x \in \Csig$, where
  $\nabla g(x)$ is the Jacobian of $g$ evaluated at $x$ and
  \item for some $L
  \geq 1$,   
  \begin{equation*}
    \frac{1}{L}\norm{x - y} \leq \norm{g(x) - g(y)} \leq L \norm{x - y} ~
    \text{for all $x,y \in \mathcal{K}$}. 
  \end{equation*}
  \end{enumerate}
  Succinctly, we assume that $\Csig$ is the image of a convex set with finite diameter 
  under a  measure preserving, bi-Lipschitz transformation.

\item
  \label{asmp: bounded_volume}
  \emph{Bounded volume:}
  We assume that neighborhood graph radius $0 < r \leq \sigma/2d$ is chosen such that
  \begin{equation*}
    2 \int_{\Csig} \Pbb(B(x,r)) f(x) dx \leq \int_{\Rd} \Pbb(B(x,r)) f(x) dx,
  \end{equation*}
  where $B(x,r)$ is the closed ball of radius $r$ at $x$.
\end{enumerate}
To motivate these conditions, we provide a brief high-level sketch
of our analysis (we return to this more formally in Section~\ref{sec: analysis}).
\citet{zhu2013} show for an arbitrary graph $G =
(V,E)$ and subset of vertices $S \subseteq V$, the PPR estimate \smash{$\Cest$}
of subset $S$ satisfies, for a constant $c>0$,
\begin{equation}
\label{eqn: graph_symmetric_set_difference_1}
\vol(\Cest \setminus S; G) + \vol(S \setminus \Cest; G) \leq c \bigl( \Phi(S,G)
\cdot \tau_{\infty}(G[S])\bigr) \vol(S;G),
\end{equation}
where $\Phi(S;G)$ is the normalized cut of $S$ (as defined in \eqref{eqn:
  normalized_cut}), and \smash{$\tau_{\infty}(G[S])$} is  the \emph{mixing
  time} of a random walk over the induced subgraph $G[S]$ (to be defined
precisely later, in \eqref{eqn: mixing_time}).  The left-hand side in
\eqref{eqn:  graph_symmetric_set_difference_1} resembles a (degree-weighted) 
form of the symmetric set difference metric in \eqref{eqn:
  misclassification_rate}, and our main goal will be to upper bound 
  this quantity. 
  Towards this goal, as we will show in Section \ref{sec:
  analysis}, the conditions \ref{asmp: bounded_density}--\ref{asmp:
  bounded_volume} allow us to upper bound the normalized cut 
\smash{$\Phi(\Csig[\Xbf]; G_{n,r})$}, and the mixing time 
\smash{$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}. In more detail, assumptions  
\ref{asmp: cluster_separation}, \ref{asmp: low_noise_density} yield an upper
bound on \smash{$\cut(\Csig[\Xbf]; G_{n,r})$}, and \ref{asmp: bounded_density}
yields a lower bound on \smash{$\vol(\Csig[\Xbf]; G_{n,r})$}; together with 
\ref{asmp: bounded_volume}, this gives an upper bound on the normalized cut.  On 
the other hand, \ref{asmp: bounded_density}, \ref{asmp: embedding} preclude
bottlenecks in the induced subgraph \smash{$G_{n,r}[\Csig[\Xbf]]$}, and combined
with the upper bound on diameter in \ref{asmp: embedding}, this leads to an upper bound on
the mixing time over this subgraph.

\subsubsection{Condition number} 
Motivated by \eqref{eqn:
  graph_symmetric_set_difference_1}, we will define $\kappa(\Cset)$ to be an
upper bound on the product \smash{$\Phi(\Csig[\Xbf]; G_{n,r}) \cdot
  \tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}. Following, the analysis of \citet{zhu2013}
  we see that the smaller $\kappa(\Cset)$ is, the
more success PPR will have in recovering the target cluster $\Cset$. We let $\theta := (r, \sigma,
\lambda, \lambda_{\sigma}, \Lambda_{\sigma}, \gamma, \diam, L)$ contain the
geometric parameters from Assumptions~\ref{asmp:  bounded_density}--\ref{asmp:
  bounded_volume}.

\begin{definition}[Well-conditioned density clusters]
  For $\lambda > 0$ and \smash{$\Cset \in \Cbb_f(\lambda)$}, let $\Cset$ satisfy 
  \ref{asmp: bounded_density}--\ref{asmp: bounded_volume} for some
  $\theta$. Then, for universal constants $c_1, c_2, c_3 > 0$ to be specified
  later, we set 
  \begin{equation}
    \label{eqn: condition_number_1}
    \Phibf(\theta) 
    := c_1 r \frac{d}{\sigma} \frac{\lambda}{\lambda_{\sigma}}
    \frac{(\lambda_{\sigma} - c_0 \frac{r^{\gamma}}{\gamma +
        1})}{\lambda_{\sigma}},~  
    \taubf(\theta) := c_2 \frac{\Lambda_{\sigma}^4 d^3 \rho^2
      L^2}{\lambda_{\sigma}^4 r^2} \log^2\left(\frac{1}{r}\right) + c_3,
  \end{equation}
  and letting \smash{$\kappa(\Cset) := \Phi_{u}(\theta) \cdot
    \tau_{u}(\theta)$}, we call $\Cset$ a 
  \emph{$\kappa$-well-conditioned density cluster}.  
\end{definition}
\noindent The condition number 
$\kappa(\Cset)$ succinctly captures the role of the various
geometric parameters. 
We note in passing that $\Phibf(\theta)$ and $\taubf(\theta)$ are exactly the upper bounds 
on \smash{$\Phi(\Csig[\Xbf]; G_{n,r})$} and
\smash{$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$} that we derive in our analysis
later, in Section \ref{sec: analysis}.

\subsubsection{Well-initialized algorithm} As is typical in the local clustering
literature, our algorithmic results will be stated with respect to specific
ranges of each of the user-specified parameters. In particular, for a
well-conditioned density cluster $\Cset$ (with respect to some $\theta$), we
require that the tuning parameters of Algorithm~\ref{alg: ppr} are chosen as
\begin{equation}
\begin{gathered}
\label{eqn: initialization}
0 < r \leq \frac{\sigma}{2d}, \quad 
\alpha \in [1/10, 1/9] \cdot \frac{1}{\taubf(\theta)}, \\
v \in \Csig[\Xbf]^g, \quad
\vol_0 \in [3/4,5/4] \cdot n(n-1) \int_{\Csig} \Pbb(B(x,r)) f(x) dx, 
\end{gathered}
\end{equation}
where \smash{$\Csig[\Xbf]^g \subseteq \Csig[\Xbf]$} will be some large
(``good'') subset of $\Csig[\Xbf]$. In particular, abbreviating $\vol_{n,r}(S)
:= \vol(S; G_{n,r})$ for $S \subseteq \Xbf$, this good subset has large volume, i.e. 
$\vol_{n,r}(\Csig[\Xbf]^g) \geq \vol_{n,r}(\Csig[\Xbf])/2$.

\begin{definition}
If the input parameters to Algorithm \ref{alg: ppr} satisfy \eqref{eqn:
  initialization} for some well-conditioned density cluster $\Cset$, we
say the algorithm is \emph{well-initialized}. 
\end{definition}
\noindent In practice it is  not feasible to set hyperparameters based on the
underlying (unknown) density $f$. Typically, one tunes PPR over a range of
hyperparameters and selects the cluster which has the smallest normalized cut. 
%it is not obvious how this scheme would affect the performance of PPR in the
%density clustering context.

>>>>

	\end{center}

\section{Examples}
\label{sec:examples}

In this section, we demonstrate the implications of our theory, and compare the performance of PPR density clustering to some alternatives, by walking through a series of examples. In all of these examples, the generative model under consideration will be a mixture of uniform distributions, plus some noise distribution. To formally define this model, we specify a support $\mathcal{X}$ and disjoint subsets $\mathcal{C}_1,\ldots,\mathcal{C}_M$ of $\mathcal{X}$, where $M \geq 2$ is the number of mixture components. We sample the background noise uniformly over $\mathcal{C}_0 = \mathcal{X} \setminus \bigcup_{m = 1}^{M} \mathcal{C}_m$, and let $\epsilon \in [0,1]$ be the fraction of background noise. The mixture distribution $\Pbb$ is therefore
\begin{equation*}
\Pbb = \frac{1 - \epsilon}{M} \sum_{m = 1}^{M} \Psi_m + \epsilon \Psi_0,
\end{equation*}
where $\Psi_m$ is the uniform distribution over $\Cset_m$.

Note that when $\epsilon$ is sufficiently small, for an appropriate choice of threshold $\lambda$ the density clustering will be $\Cbb_f(\lambda) = \set{\Cset_1,\ldots,\Cset_M}$. In this setup, therefore, estimating a density cluster is equivalent to recovering a mixture component.
\subsection{1d uniform example.}

We show the performance of our bounds in a mixture of uniforms in $\Reals$, and compare with the bound obtained via a global spectral clustering approach.

\subsection{General uniform example.}

We show the performance of our bounds in a mixture of uniforms in $\Reals^d$, and compare with the bound obtained via a density estimation approach.

\subsection{Lower bound.}

The theory we have developed so far gives tighter guarantees for plug-in density cluster estimators than for PPR. However, in both cases these are merely upper bounds on the clustering error. Now, we provide accompanying lower bounds for the PPR algorithm. These will demonstrate that there exist densities for which consistent estimation of density clusters is still possible, but for which \emph{PPR} will provably fail to recover density clusters (with high probability.) Moreover, they will show that the difficulty function $\kappa(\Cset)$ exhibits the right dependence on (some of) the geometric parameters we have introduced.

Our hard case will be a mixture of $2$-uniforms with background noise. Given $0 < \sigma < \rho$, let $\mathcal{X} = [-1.5\sigma,1.5\sigma] \times [-.5\rho,.5\rho]$, and let $\Cset_1 = [-1.5\sigma,-.5\sigma] \times [.5\rho,1.5\rho]$, $\Cset_2 = [.5\sigma,1.5\sigma] \times [-.5\rho,.5\rho]$, so that the background noise is sampled over $\Cset_0 = [-.5\sigma,.5\sigma] \times [-.5\rho,.5\rho]$. We denote the resulting model by $\Pbb(\sigma,\rho,\epsilon)$ to emphasize the dependence of the distribution on the parameters of interest.

\section{Technical results.}

\subsection{Lower bound.}

Let $\Sset = [-1.5\sigma,1.5\sigma] \times [-\rho/2,0]$. Let the \emph{cut} and \emph{volume} functionals be defined for $\Aset \subset \Rd$ as follows:
\begin{equation*}
\cut_{\Pbb,r}(\Aset) := \int_{\Aset} \int_{\mathcal{X} \setminus \Aset} \1(\norm{x - y} \leq r) \,d\Pbb(x) \,d\Pbb(y), \vol_{\Pbb,r}(\Aset) := \int_{\Aset} \int_{\mathcal{X}} \1(\norm{x - y} \leq r) \,d\Pbb(x) \,d\Pbb(y)
\end{equation*}
and the \emph{normalized cut},
\begin{equation*}
\Phi_{\Pbb,r}(\Aset) := \frac{\cut_{\Pbb,r}(\Aset)}{\min\{\vol_{\Pbb,r}(\Aset),\vol_{\Pbb,r}(\mathcal{X} \setminus \Aset)\}}.
\end{equation*}

\textbf{1.} We first upper bound the normalized cut $\Phi_{\Pbb,r}(\Sset)$ and lower bound the normalized cut $\Phi_{\Pbb,r}(\Cset_1)$, using the geometric parameters $\sigma$ and $\rho$, the \textcolor{red}{drop parameter} $\epsilon$, and the radius $r$. Our bounds will be crude, but will display the right dependence on these parameters, and so will be sufficient for our purposes.

We begin with the upper bound on $\Phi_{\Pbb,r}(\Sset)$. Note that only points $x = (x_1,x_2)$ such that $-r < x_2 < 0$ contribute to the cut of $\Sset$. Letting $f$ be the density function of $\Pbb$, we have
\begin{align*}
\cut_{\Pbb,r}(\Sset) & \leq \Pbb(\set{x \in \mathcal{X}: -r < x_2 < 0}) \cdot \max_{x \in \mathcal{X}} \Pbb(B(x,r)) \\
& \leq \frac{3 r}{\rho} \cdot \frac{\nu_d r^d}{\sigma \rho}.
\end{align*}
By symmetry, $\vol_{\Pbb,r}(\Sset) = \vol_{\Pbb,r}(\mathcal{X} \setminus \Sset)$, and therefore it is sufficient to lower bound $\vol_{\Pbb,r}(\Sset)$. We have
\begin{align*}
\vol_{\Pbb,r}(\Sset) & \geq \Pbb(\set{x \in \Cset_1 \cap \Sset: \dist(x, \partial \Cset_1) > r}) \cdot \frac{\nu_d r^d}{\sigma \rho} \\
& = \frac{(\sigma - 2r)(\rho - r)}{2 \sigma \rho} \cdot \frac{\nu_d r^d}{\sigma \rho}  \\
& \geq \frac{3\nu_d r^d}{16\sigma \rho},
\end{align*}
where the last inequality follows since $r \leq \sigma/4$ and $\sigma < \rho$. Therefore,
\begin{equation*}
\Phi_{\Pbb,r}(\Sset) \leq 16 \frac{r}{\rho}.
\end{equation*}
Similar manipulations yield a lower bound on $\Phi_{\Pbb,r}(\Cset_1)$. We use our estimates of the volume of spherical caps from \textcolor{red}{A.6} to obtain a lower bound on $\cut_{\Pbb,r}(\Cset_1)$, 
\begin{align*}
\cut_{\Pbb,r}(\Cset_1) & \geq \Pbb(\set{x \in \Cset_1: \dist(x,\Cset_0) < r/32}) \cdot \frac{\nu(\textrm{cap}_{r}(\frac{31r}{32})) \epsilon}{2\rho\sigma} \\
& \geq \frac{r (1 - \epsilon)}{32 \sigma} \cdot \frac{3 \nu_d r^d \epsilon}{16\rho\sigma}.
\end{align*}
and we can immediately upper bound the volume of $\Cset_1$,
\begin{equation*}
\vol_{\Pbb,r}(\Cset_1) \leq (1 - \epsilon) \cdot  \frac{(1 - \epsilon)}{\rho \sigma} \nu_d r^d.
\end{equation*}
Observe that $\vol_{\Pbb,r}(\mathcal{X} \setminus \Cset_1) \geq \vol_{\Pbb,r}(\Cset_2) = \vol_{\Pbb,r}(\Cset_1)$. Therefore,
\begin{equation}
\label{eqn:lb_proof_2}
\Phi_{\Pbb,r}(\Cset_1) \geq  \frac{\epsilon r}{256 \sigma}.
\end{equation}

\textbf{2.} The upper bound \eqref{eqn:lb_proof_2} can be extended to hold for all sets $\Aset$ with small symmetric set difference $\Aset \triangle \Cset_1$. Formally,
\begin{equation}
\label{eqn:continuity_normalized_cut}
\Phi_{\Pbb,r}(\Aset) \geq \frac{1}{2\pi} \cdot \frac{\epsilon^2r}{\sigma} \bigl(1 - 4 \cdot \Pbb(\Aset \triangle \Cset_1 )\bigr)
\end{equation}
The proof of \eqref{eqn:continuity_normalized_cut} is somewhat involved, and we defer it to the following section.

\textbf{3.} We now convert from population to sample normalized cut. \textit{Claim 1:} There exist constants $c_1,c_2$ which may depend on $(\epsilon, \rho, \sigma)$ such that the following statements holds with respect to $c_1,c_2$: For all $n$ such that
\begin{equation*}
c_1 \cdot \left(\frac{\log n}{n}\right)^{1/d} \leq r,
\end{equation*}
the sample normalized cut of $\Aset$ is lower bounded
\begin{equation*}
\Phi_{n,r}(\Aset) \geq \frac{\Phi_{\Pbb,r}(\Aset)}{4} \quad \textrm{for all $\Aset \subset \Rd$,}
\end{equation*}  
and the sample normalized cut of $\Sset$ is upper bounded,
\begin{equation*}
\Phi_{n,r}(\Sset) \leq \frac{\Phi_{\Pbb,r}(\Sset)}{2}
\end{equation*}
with probability at least $1 - \frac{c_2}{n}$.

\textbf{4.}

We prove this claim in a following section.

\subsubsection{Proof of \eqref{eqn:continuity_normalized_cut}}
To prove \eqref{eqn:continuity_normalized_cut}, we will need to upper bound the cut of $\Aset$, and lower bound the volume. We begin by upper bounding the cut, which will be the more difficult of the two. To obtain the desired result, we will need to discretize $\Aset$. For simplicity, without loss of generality assume $\rho$ and $\sigma$ are constant multiples of $r$. For $i = 0,\ldots,\frac{6\sigma}{r} - 1$ and $j = 0,\ldots, \frac{2\rho}{r} - 1$, let
